---
title: "Case-Crossover Simulation Demo"
author: "Ziang Zhang"
date: "2026-04-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Goal

This file shows a minimal workflow for:

1. simulating `dailymort_pulm_nonsenior` with `simulate_casecrossover_data()`;
2. fitting a case-crossover model with four simple smooth effects using `BayesGP`;
3. extracting the posterior for each effect function with `predict()`;
4. simulating and fitting a simpler PM2.5-only data-generating process.

The simulation below uses deliberately high-contrast effects and sets the total
case count in each stratum explicitly before the multinomial redistribution.
The fit uses all strata under the grouped multinomial case-crossover likelihood.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

source("analysis/simulate_casecrossover_data.R")

library(BayesGP)
```


Define small helpers for extracting and plotting posterior effects. Because the
model uses `initial_location = pm25_reference` and `family = "cc"` has no
intercept, `predict()` should already return an effect that is anchored at the
reference value.

```{r posterior-helpers}
posterior_effect <- function(fit, data, variable, true_fun, n_grid = 50,
                             reference = NULL) {
  variable_range <- range(data[[variable]], na.rm = TRUE)
  grid_values <- seq(
    variable_range[1],
    variable_range[2],
    length.out = n_grid
  )

  if (!is.null(reference)) {
    if (!is.numeric(reference) || length(reference) != 1 || !is.finite(reference)) {
      stop("`reference` must be NULL or a finite numeric scalar.", call. = FALSE)
    }
    if (reference < variable_range[1] || reference > variable_range[2]) {
      stop("`reference` must lie inside the observed range of `variable`.", call. = FALSE)
    }
    grid_values <- sort(unique(c(grid_values, reference)))
  }

  newdata <- data.frame(grid_values)
  names(newdata) <- variable

  summary_df <- predict(
    fit,
    newdata = newdata,
    variable = variable,
    only.samples = FALSE
  )

  summary_df$true <- true_fun(summary_df[[variable]])
  summary_df
}

plot_effect <- function(effect_df, variable, x_label) {
  x <- effect_df[[variable]]
  y_range <- range(
    effect_df$q0.025,
    effect_df$q0.975,
    effect_df$true,
    finite = TRUE
  )

  plot(
    x,
    effect_df$mean,
    type = "n",
    ylim = y_range,
    xlab = x_label,
  ylab = "Centered effect",
    main = x_label
  )

  polygon(
    c(x, rev(x)),
    c(effect_df$q0.025, rev(effect_df$q0.975)),
    col = grDevices::adjustcolor("grey70", alpha.f = 0.45),
    border = NA
  )
  lines(x, effect_df$mean, lwd = 2)
  lines(x, effect_df$true, col = "red", lwd = 2, lty = 2)
  legend(
    "topleft",
    legend = c("Posterior mean", "True function"),
    lty = c(1, 2),
    lwd = 2,
    col = c("black", "red"),
    bty = "n"
  )
}
```


## Simulate A Toy Case-Crossover Dataset

```{r simulate-data}
data_env <- new.env(parent = emptyenv())
load("data/data_CD3518_sample_2.RData", envir = data_env)
base_data <- data_env$data_CD3518_sample_2

simulation_base_data <- base_data
simulation_base_data$strata <- format(simulation_base_data$Date, "%Y%b%A")
cases_per_stratum <- 50L
pm25_reference <- 5
f_pm25_true <- function(x) {
  0.75 * (log1p(x) - log1p(pm25_reference))
}

true_deriv <- 0.75 * 1/(pm25_reference + 1)
```


## PM2.5-Only Diagnostic

The simulation helper also accepts `NULL` effects. Any omitted effect is treated
as zero in the true linear predictor. This makes it easy to create a one-function
data-generating process and fit the matching one-function model.

```{r pm25-only-simulation}
pm25_only_data <- simulate_casecrossover_data(
  f_pm25 = f_pm25_true,
  data = simulation_base_data,
  seed = 20260425,
  stratum_total_cases = cases_per_stratum
)

pm25_only_totals <- tapply(
  pm25_only_data$dailymort_pulm_nonsenior,
  pm25_only_data$strata,
  sum
)

stopifnot(all(pm25_only_totals == cases_per_stratum))

pm25_only_fit_data <- as.data.frame(
  pm25_only_data[order(pm25_only_data$strata, pm25_only_data$Date), ]
)

data.frame(
  rows = nrow(pm25_only_fit_data),
  strata = length(unique(pm25_only_fit_data$strata)),
  total_cases = sum(pm25_only_fit_data$dailymort_pulm_nonsenior),
  min_cases_per_stratum = min(pm25_only_totals),
  max_cases_per_stratum = max(pm25_only_totals)
)
```

Fit only the PM2.5 smooth term:

```{r pm25-only-fit}
pm25_shift <- pm25_reference

fit_pm25_only <- model_fit(
  formula = dailymort_pulm_nonsenior ~
    f(
      pm25Sample,
      model = "tiwp2",
      computation = "fem",
      initial_location = pm25_reference,
      a = 2, # parameter for mgp or tiwp2 only, not for other models
      # boundary.prior = list(mean = true_deriv, prec = 1e10),
      normalized_boundary = TRUE
    ),
  family = "cc",
  strata = "strata",
  weight = NULL,
  data = pm25_only_fit_data,
  method = "aghq",
  aghq_k = 3
)

plot(fit_pm25_only)
est_derivative <- mean(extract_boundary_samples(fit_pm25_only, "pm25Sample"))

print(
  data.frame(
    true_derivative = true_deriv,
    estimated_derivative = est_derivative
  )
)
```

Extract and plot the PM2.5 posterior curve from the simpler model:

```{r pm25-only-posterior, fig.width=6, fig.height=4}
pm25_only_posterior <- posterior_effect(
  fit = fit_pm25_only,
  data = pm25_only_fit_data,
  variable = "pm25Sample",
  true_fun = f_pm25_true,
  reference = pm25_reference
)

plot_effect(pm25_only_posterior, "pm25Sample", "PM2.5 only")
```
