---
title: "casecross"
author: "Ziang Zhang"
date: "2024-08-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r}
load("data/data_CD3518_sample_2.RData")
head(data_CD3518_sample_2)
data_CD3518_sample_2$strata = format(data_CD3518_sample_2$Date, '%Y%b%A')

NperStrata = tapply(data_CD3518_sample_2$dailymort_pulm_nonsenior, data_CD3518_sample_2$strata, sum)
haveData = which(NperStrata > 0)

datSub = as.data.frame(data_CD3518_sample_2[data_CD3518_sample_2$strata %in% names(haveData), ])

datSub$dateFac = factor(as.character(datSub$Date))
datSub$dateInt = as.integer(datSub$dateFac)
```


## Fitting the data using mGP:

```{r}
library(tidyverse)
library(Matrix)
library(knitr)
devtools::load_all("BayesGP", quiet = TRUE)
```

In BayesGP, `c` is specified on the original PM2.5 scale, so with `a = 2` the
base function is `sqrt(pm25Sample + c)`. We use `c = 0` so that the base
function is `sqrt(pm25Sample)`, and place the PSD prior reference at the
smallest observed PM2.5 value.

```{r}
datSub <- datSub %>% arrange(pm25Sample)
pm25_min <- min(datSub$pm25Sample)
```

The fastest way to fit the model is to use the FEM approximation:

```{r}
mod_mgp <- model_fit(
  formula = dailymort_pulm_nonsenior ~
    f(
      pm25Sample,
      model = "mgp",
      computation = "fem",
      a = 2,
      c = 0,
      k = 50,
      region = pm25_min + c(0, 40),
      normalized_boundary = TRUE,
      boundary.prior = list(mean = 0, prec = 0.001),
      sd.prior = list(prior = "exp", param = list(u = 0.01, alpha = 0.5), h = 5, x = pm25_min)
    ),
  family = "cc",
  strata = "strata",
  weight = NULL,
  data = datSub,
  M = 3000,
  method = "aghq"
)

plot(mod_mgp)
```


Alternatively, one can also use the state-space representation, which is exact but computationally less stable:

```{r}
# round the x values to 1 decimal place to make the computation more stable
datSub$pm25Sample_rounded <- round(datSub$pm25Sample, 1)
mod_mgp2 <- model_fit(
  formula = dailymort_pulm_nonsenior ~
    f(
      pm25Sample_rounded,
      model = "mgp",
      computation = "state-space",
      a = 2,
      c = 0,
      normalized_boundary = TRUE,
      boundary.prior = list(mean = 0, prec = 0.001),
      sd.prior = list(prior = "exp", param = list(u = 0.01, alpha = 0.5), h = 5, x = pm25_min)
    ),
  family = "cc",
  strata = "strata",
  weight = NULL,
  data = datSub,
  M = 3000,
  method = "aghq"
)

plot(mod_mgp2)
```

## Fitting the data using tIWP2:

Using the FEM approximation:

```{r}
mod_tiwp2 <- model_fit(
  formula = dailymort_pulm_nonsenior ~
    f(
      pm25Sample,
      model = "tiwp2",
      computation = "fem",
      a = 2,
      c = 0,
      k = 50,
      region = pm25_min + c(0, 40),
      normalized_boundary = TRUE,
      boundary.prior = list(mean = 0, prec = 0.001),
      sd.prior = list(prior = "exp", param = list(u = 0.01, alpha = 0.5), h = 5, x = pm25_min)
    ),
  family = "cc",
  strata = "strata",
  weight = NULL,
  data = datSub,
  M = 3000,
  method = "aghq"
)

plot(mod_tiwp2)
```

Using the state-space representation, again on the rounded PM2.5 values:

```{r}
mod_tiwp2_ss <- model_fit(
  formula = dailymort_pulm_nonsenior ~
    f(
      pm25Sample_rounded,
      model = "tiwp2",
      computation = "state-space",
      a = 2,
      c = 0,
      normalized_boundary = TRUE,
      boundary.prior = list(mean = 0, prec = 0.001),
      sd.prior = list(prior = "exp", param = list(u = 0.01, alpha = 0.5), h = 5, x = pm25_min)
    ),
  family = "cc",
  strata = "strata",
  weight = NULL,
  data = datSub,
  M = 3000,
  method = "aghq"
)

plot(mod_tiwp2_ss)
```
