---
title: "Illustration"
author: "Ziang Zhang"
date: "2024-08-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this example, I will illustrate how to fit the tIWP2 or mGP method, induced by different monotone transformations.

Let's load the relevant functions and libraries that we will be using in this example:
```{r}
library(tidyverse)
library(Matrix)
devtools::load_all("BayesGP", quiet = TRUE)
```

First, let's simulate the following dataset

Simulate some random points:
```{r}
obs_max <- 8
samp_max <- 10
c <- 1
sd_noise <- 4
n <- 100
set.seed(123)
f <- function(x) (3+4*log(x + c))^1.5
x <- seq(0, samp_max, length.out = (n+1))[-1]
y <- f(x) + rnorm(n, sd = sd_noise)
plot(x, y, ylab = "y", xlab = "x")
lines(x, f(x), col = "red")
```

To make the experiment a bit more interesting, let's consider some training and testing separation:
```{r}
data_sim <- data.frame(x = x, y = y) %>% arrange(x)
data_train <- data_sim %>% filter(x < obs_max)
```


The true function $f$ is close (but not exactly equal) to some logarithm function.


## Fitting the Exact Process

First, we consider fitting the exact process through the augmented state-space representation. The retained notebooks now load the in-repo `BayesGP` fork directly and use a unified formula interface for `iwp`, `tiwp2`, and `mgp`, instead of sourcing helper scripts and compiling a separate TMB template by hand.

For now, let's assume the mGP and tIWP are both induced by the monotone function $m(x) = \sqrt{x+c}$.


For comparison purpose, let's assume the prior on the PSD $\sigma_8(2)$ is always an Exponential with median at $5$.

### IWP2

Fitting the IWP2:

```{r}
prior_median <- 5
iwp2 <- model_fit(
  formula = y ~ f(
    x,
    model = "iwp",
    computation = "state-space",
    order = 2,
    initial_location = "left",
    grid = data_sim$x,
    sd.prior = list(prior = "exp", param = list(u = prior_median, alpha = 0.5), h = 2)
  ),
  data = data_train,
  family = "gaussian",
  control.family = list(sd = sd_noise),
  M = 3000
)
samples_iwp2 <- as.matrix(
  predict(iwp2, newdata = data_sim, variable = "x", only.samples = TRUE)[, -1, drop = FALSE]
)

iwp_summary <- data.frame(x = data_sim$x, mean = rowMeans(samples_iwp2),
                          lower = apply(samples_iwp2, 1, quantile, probs = 0.025),
                          upper = apply(samples_iwp2, 1, quantile, probs = 0.975))
```

Visualize the IWP2:

```{r}
plot(data_sim$x, data_sim$y, type = "p", col = "black", 
     lwd = 1, pch = 1, cex = 0.2,
     ylim = c(0,100), ylab = "y", xlab = "x",
     )
lines(iwp_summary$x, iwp_summary$mean, col = "red", lwd = 2)
lines(iwp_summary$x, f(iwp_summary$x), col = "blue", lwd = 2, lty = 2)
matlines(x = iwp_summary$x, y = samples_iwp2[,2:5], lty = 1, col = "blue", lwd = 0.5)
polygon(c(iwp_summary$x, rev(iwp_summary$x)), c(iwp_summary$lower, rev(iwp_summary$upper)), col = rgb(1, 0, 0, 0.2), border = NA)
abline(v = obs_max, col = "purple", lty = 2)
```


### tIWP2

Fitting the tIWP2:

```{r}
tiwp2 <- model_fit(
  formula = y ~ f(
    x,
    model = "tiwp2",
    computation = "state-space",
    a = 2,
    c = c,
    grid = data_sim$x,
    sd.prior = list(prior = "exp", param = list(u = prior_median, alpha = 0.5), h = 2, x = 8)
  ),
  data = data_train,
  family = "gaussian",
  control.family = list(sd = sd_noise),
  M = 3000
)
samples_tiwp2 <- as.matrix(
  predict(tiwp2, newdata = data_sim, variable = "x", only.samples = TRUE)[, -1, drop = FALSE]
)
```

Visualize the tIWP2:

```{r}
tiwp_summary <- data.frame(x = data_sim$x, mean = rowMeans(samples_tiwp2),
                          lower = apply(samples_tiwp2, 1, quantile, probs = 0.025),
                          upper = apply(samples_tiwp2, 1, quantile, probs = 0.975))

plot(data_sim$x, data_sim$y, type = "p", col = "black", 
     lwd = 1, pch = 1, cex = 0.2,
     ylim = c(0,100), ylab = "y", xlab = "x")
lines(tiwp_summary$x, tiwp_summary$mean, col = "red", lwd = 2)
lines(tiwp_summary$x, f(tiwp_summary$x), col = "blue", lwd = 2, lty = 2)
matlines(x = tiwp_summary$x, y = samples_tiwp2[,2:5], lty = 1, col = "blue", lwd = 0.5)
polygon(c(tiwp_summary$x, rev(tiwp_summary$x)), c(tiwp_summary$lower, rev(tiwp_summary$upper)), col = rgb(1, 0, 0, 0.2), border = NA)
abline(v = obs_max, col = "purple", lty = 2)
```


### mGP

Fitting the mGP:

```{r}
mGP <- model_fit(
  formula = y ~ f(
    x,
    model = "mgp",
    computation = "state-space",
    a = 2,
    c = c,
    grid = data_sim$x,
    normalized_boundary = TRUE,
    sd.prior = list(prior = "exp", param = list(u = prior_median, alpha = 0.5), h = 2, x = 8)
  ),
  data = data_train,
  family = "gaussian",
  control.family = list(sd = sd_noise),
  M = 3000
)
samples_mGP <- as.matrix(
  predict(mGP, newdata = data_sim, variable = "x", only.samples = TRUE)[, -1, drop = FALSE]
)
```

```{r}
mGP_summary <- data.frame(x = data_sim$x, mean = rowMeans(samples_mGP),
                          lower = apply(samples_mGP, 1, quantile, probs = 0.025),
                          upper = apply(samples_mGP, 1, quantile, probs = 0.975))

plot(data_sim$x, data_sim$y, type = "p", col = "black", 
     lwd = 1, pch = 1, cex = 0.2,
     ylim = c(0,100), ylab = "y", xlab = "x")
lines(mGP_summary$x, mGP_summary$mean, col = "red", lwd = 2)
lines(mGP_summary$x, f(mGP_summary$x), col = "blue", lwd = 2, lty = 2)
matlines(x = mGP_summary$x, y = samples_mGP[,2:5], lty = 1, col = "blue", lwd = 0.5)
polygon(c(mGP_summary$x, rev(mGP_summary$x)), c(mGP_summary$lower, rev(mGP_summary$upper)), col = rgb(1, 0, 0, 0.2), border = NA)
abline(v = obs_max, col = "purple", lty = 2)
```


## Fitting the FEM Approximation

When the locations are densely irregularly placed over the region, the exact computation tends to become challenging, even with the state-space representation that simplifies the precision matrix.
In such scenario, it might be preferable to use a basis representation constructed from the FEM approximation.

Here we will take the mGP as an example, mirroring the published illustration page. The original workflow used `k = 30`, so we keep that setting here in order to reproduce the same approximation experiment with the compact package interface.

```{r}
k_fem <- 30
```

```{r}
mGP_FEM <- model_fit(
  formula = y ~ f(
    x,
    model = "mgp",
    computation = "fem",
    a = 2,
    c = c,
    k = k_fem,
    grid = data_sim$x,
    normalized_boundary = TRUE,
    sd.prior = list(prior = "exp", param = list(u = prior_median, alpha = 0.5), h = 2, x = 8)
  ),
  data = data_train,
  family = "gaussian",
  control.family = list(sd = sd_noise),
  M = 3000
)
```

Once the model is fitted, we can compute the posterior sample of the function values at the test locations.

```{r}
samples_mGP_FEM <- as.matrix(
  predict(mGP_FEM, newdata = data_sim, variable = "x", only.samples = TRUE)[, -1, drop = FALSE]
)
```

```{r}
mGP_FEM_summary <- data.frame(x = data_sim$x, mean = rowMeans(samples_mGP_FEM),
                          lower = apply(samples_mGP_FEM, 1, quantile, probs = 0.025),
                          upper = apply(samples_mGP_FEM, 1, quantile, probs = 0.975))

plot(data_sim$x, data_sim$y, type = "p", col = "black", 
     lwd = 1, pch = 1, cex = 0.2,
     ylim = c(0,100), ylab = "y", xlab = "x")
lines(mGP_FEM_summary$x, mGP_FEM_summary$mean, col = "red", lwd = 2)
lines(mGP_FEM_summary$x, f(mGP_FEM_summary$x), col = "blue", lwd = 2, lty = 2)
matlines(x = mGP_FEM_summary$x, y = samples_mGP_FEM[,2:5], lty = 1, col = "blue", lwd = 0.5)
polygon(c(mGP_FEM_summary$x, rev(mGP_FEM_summary$x)), c(mGP_FEM_summary$lower, rev(mGP_FEM_summary$upper)), col = rgb(1, 0, 0, 0.2), border = NA)
abline(v = obs_max, col = "purple", lty = 2)
```
