---
title: "Prior Covariance Convergence for exact and O-spline IWP2"
author: "Ziang Zhang"
date: "2026-04-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Goal

This notebook checks whether the newly added exact order-2 `iwp`
`computation = "state-space"` path in the local `BayesGP` fork agrees with the
package's native O-spline approximation at the prior covariance level.

The comparison is done on the stochastic process part only. The boundary/global
polynomial term is fixed at zero so that both constructions are compared on the
same latent object.

Because both paths scale linearly in `sd^2`, the convergence check is done at
unit scale without loss of generality.

## Setup

```{r}
library(tidyverse)
library(Matrix)
devtools::load_all("BayesGP", quiet = TRUE)
```

The exact state-space `iwp` implementation currently requires left-boundary
anchoring, so the evaluation grids are shifted to start at zero before the
covariance is constructed.

For the native O-spline approximation, `k` denotes the knot count used in the
underlying `iwp` object. This produces `k - 1` local O-spline coefficients.

```{r}
evaluation_grids <- tibble(
  grid_case = c("balanced irregular", "boundary clustered"),
  x = list(
    c(0.10, 0.35, 0.75, 1.30, 2.05, 2.95, 4.00, 5.20, 6.60, 8.15),
    c(0.20, 0.24, 0.31, 0.52, 1.00, 1.80, 3.00, 4.70, 7.00, 8.50)
  )
)

k_values <- c(10, 20, 40, 80, 160, 320)
```

## Helper Functions

The exact covariance is built from the same exact observation operator used by
the new state-space pathway:

$$
\Sigma_{\mathrm{exact}} = B_{\mathrm{exact}} Q_{\mathrm{exact}}^{-1} B_{\mathrm{exact}}^\top.
$$

The native O-spline covariance is built from the original BayesGP machinery:

$$
\Sigma_{\mathrm{ospline}} = B_{\mathrm{ospline}} Q_{\mathrm{ospline}}^{-1} B_{\mathrm{ospline}}^\top.
$$

```{r}
build_exact_iwp_covariance <- function(x){
  shifted_x <- sort(x - min(x))
  support_x <- shifted_x[shifted_x > 0]

  if(length(support_x) == 0){
    return(matrix(0, nrow = length(shifted_x), ncol = length(shifted_x)))
  }

  exact_B <- build_exact_observation_matrix(
    observed_x = shifted_x,
    support_x = support_x
  )
  exact_P <- as.matrix(IWP_joint_prec(t_vec = support_x, p = 2))

  as.matrix(exact_B %*% solve(exact_P, t(exact_B)))
}

build_ospline_iwp_covariance <- function(x, k){
  shifted_x <- sort(x - min(x))
  knot_sequence <- seq(0, max(shifted_x), length.out = k)
  temp_data <- data.frame(z = shifted_x)

  temp_instance <- methods::new(
    "iwp",
    response_var = as.name("y"),
    smoothing_var = as.name("z"),
    order = 2,
    knots = knot_sequence,
    k = k,
    observed_x = shifted_x,
    sd.prior = list(),
    psd.prior = list(),
    boundary.prior = list(),
    data = temp_data,
    X = matrix(0, nrow = 0, ncol = 0),
    B = matrix(0, nrow = 0, ncol = 0),
    P = matrix(0, nrow = 0, ncol = 0),
    initial_location = 0,
    region = range(c(0, shifted_x))
  )

  ospline_B <- as.matrix(local_poly(temp_instance))
  ospline_P <- as.matrix(compute_weights_precision(temp_instance))

  as.matrix(ospline_B %*% solve(ospline_P, t(ospline_B)))
}

frobenius_relative_error <- function(approximate_cov, exact_cov){
  sqrt(sum((approximate_cov - exact_cov)^2) / sum(exact_cov^2))
}

max_relative_error <- function(approximate_cov, exact_cov){
  max(abs(approximate_cov - exact_cov)) / max(abs(exact_cov))
}

variance_relative_error <- function(approximate_cov, exact_cov){
  max(abs(diag(approximate_cov) - diag(exact_cov))) / max(abs(diag(exact_cov)))
}

format_results_table <- function(table_df){
  table_df %>%
    mutate(
      k = as.integer(k),
      `Basis coefficients` = k - 1L,
      `Relative Frobenius error` = signif(frobenius_rel_error, 6),
      `Max relative error` = signif(max_rel_error, 6),
      `Variance relative error` = signif(variance_rel_error, 6)
    ) %>%
    select(
      `Grid case` = grid_case,
      k,
      `Basis coefficients`,
      `Relative Frobenius error`,
      `Max relative error`,
      `Variance relative error`
    )
}
```

## Run the Comparison

```{r}
comparison_results <- evaluation_grids %>%
  tidyr::crossing(k = k_values) %>%
  mutate(
    exact_covariance = map(x, build_exact_iwp_covariance),
    ospline_covariance = map2(x, k, build_ospline_iwp_covariance),
    frobenius_rel_error = map2_dbl(
      ospline_covariance,
      exact_covariance,
      frobenius_relative_error
    ),
    max_rel_error = map2_dbl(
      ospline_covariance,
      exact_covariance,
      max_relative_error
    ),
    variance_rel_error = map2_dbl(
      ospline_covariance,
      exact_covariance,
      variance_relative_error
    )
  ) %>%
  select(grid_case, k, frobenius_rel_error, max_rel_error, variance_rel_error)

knitr::kable(
  format_results_table(comparison_results),
  caption = "Exact order-2 IWP state-space covariance versus native O-spline covariance.",
  align = c("l", "r", "r", "r", "r", "r")
)
```

```{r}
dir.create("output", showWarnings = FALSE)
write.csv(
  comparison_results,
  file = "output/iwp_prior_covariance_convergence_metrics.csv",
  row.names = FALSE
)
```

## Error Trends

```{r}
comparison_results %>%
  ggplot(aes(
    x = k,
    y = frobenius_rel_error,
    color = grid_case,
    group = grid_case,
    linetype = grid_case,
    shape = grid_case
  )) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.2) +
  scale_x_log10(breaks = k_values) +
  scale_y_log10() +
  labs(
    x = "Number of O-spline knots (k)",
    y = "Relative Frobenius error",
    color = "Grid",
    linetype = "Grid",
    shape = "Grid",
    title = "Native O-spline IWP covariance approaches the exact state-space covariance"
  ) +
  theme_bw()
```

```{r}
comparison_results %>%
  ggplot(aes(
    x = k,
    y = max_rel_error,
    color = grid_case,
    group = grid_case,
    linetype = grid_case,
    shape = grid_case
  )) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.2) +
  scale_x_log10(breaks = k_values) +
  scale_y_log10() +
  labs(
    x = "Number of O-spline knots (k)",
    y = "Maximum relative entrywise error",
    color = "Grid",
    linetype = "Grid",
    shape = "Grid",
    title = "Entrywise covariance error also shrinks under O-spline refinement"
  ) +
  theme_bw()
```

## Largest-k Summary

```{r}
largest_k_summary <- comparison_results %>%
  group_by(grid_case) %>%
  filter(k == max(k)) %>%
  ungroup()

knitr::kable(
  format_results_table(largest_k_summary),
  caption = "Largest-k summary for the IWP covariance comparison.",
  align = c("l", "r", "r", "r", "r", "r")
)
```

## Takeaway

For the exact order-2 `iwp` state-space path added in this local fork, the
native BayesGP O-spline approximation is numerically consistent at the prior
covariance level:

- the covariance error decreases rapidly as `k` grows,
- the decrease is stable across two irregular evaluation grids,
- and the refined O-spline covariance is extremely close to the exact
  state-space covariance.

So this check supports the claim that the new exact state-space implementation
is coherent with the package's original O-spline approximation for `iwp`
when the two are compared under matched left-boundary anchoring.
