---
title: "Prior Covariance Convergence for mGP and tIWP2"
author: "Ziang Zhang"
date: "2026-04-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Goal

This notebook checks whether the FEM approximations for `mGP` and `tIWP2`
converge to the exact prior covariance implied by the exact state-space
construction.

The comparison focuses on the stochastic process covariance only. In other
words, the boundary/global polynomial term is held fixed at zero so that the
exact and FEM paths are compared on the same latent process object.

## Setup

```{r}
library(tidyverse)
library(Matrix)
devtools::load_all("BayesGP", quiet = TRUE)
```

We use an irregular evaluation grid and vary both the curvature parameter `a`
and the shift parameter `c`.

For `mGP`, increasing the FEM basis size alone is not the full refinement path,
because `Compute_Prec()` itself is built through numerical integration. So the
`mGP` experiment refines both:

- the FEM basis size `k`,
- and the integration step size `accuracy`.

For `tIWP2`, the FEM path is an O-spline construction on the transformed scale,
so `k` alone is the relevant refinement parameter here.

```{r}
evaluation_grid <- c(0.25, 0.45, 0.70, 1.05, 1.40, 1.80, 2.30, 2.90, 3.50, 4.20)
k_values <- c(20, 40, 80, 160)
parameter_grid <- expand.grid(
  model = c("mgp", "tiwp2"),
  a = c(0.5, 1, 2, -1),
  c_shift = c(0.5, 1.5),
  stringsAsFactors = FALSE
) %>%
  as_tibble()
```

## Helper Functions

The exact covariance is built by inverting the exact state-space precision and
extracting the function-value block.

The FEM covariance is built directly from the FEM design matrix and precision
matrix:

$$
\Sigma_{\mathrm{FEM}} = B Q^{-1} B^\top.
$$

```{r}
extract_function_block <- function(covariance_matrix){
  covariance_matrix[
    seq(1, nrow(covariance_matrix), by = 2),
    seq(1, ncol(covariance_matrix), by = 2),
    drop = FALSE
  ]
}

build_exact_prior_covariance <- function(model_name, x, a, c_shift){
  if(model_name == "mgp"){
    precision <- as.matrix(mGP_joint_prec(t_vec = x, a = a, c = c_shift))
    return(extract_function_block(solve(precision)))
  }

  if(model_name == "tiwp2"){
    transform <- make_boxcox_transform(a = a, c = c_shift, normalized = TRUE, ref_location = 0)
    transformed_x <- transform(x)
    precision <- as.matrix(IWP_joint_prec(t_vec = transformed_x, p = 2))
    return(extract_function_block(solve(precision)))
  }

  stop("Unsupported model.")
}

build_fem_prior_covariance <- function(model_name, x, a, c_shift, k,
                                       mgp_accuracy = 0.002){
  if(model_name == "mgp"){
    region <- range(c(0, x))
    design <- as.matrix(
      Compute_Design(x = x, k = k, region = region, boundary = TRUE)
    )
    precision <- as.matrix(
      Compute_Prec(
        a = a,
        c = c_shift,
        k = k,
        region = region,
        accuracy = mgp_accuracy,
        boundary = TRUE
      )
    )
    return(design %*% solve(precision, t(design)))
  }

  if(model_name == "tiwp2"){
    transform <- make_boxcox_transform(a = a, c = c_shift, normalized = TRUE, ref_location = 0)
    transformed_x <- transform(x)
    transformed_knots <- seq(0, max(transformed_x), length.out = k)
    transformed_data <- data.frame(z = transformed_x)

    temp_instance <- methods::new(
      "iwp",
      response_var = as.name("y"),
      smoothing_var = as.name("z"),
      order = 2,
      knots = transformed_knots,
      k = k,
      observed_x = sort(transformed_x),
      sd.prior = list(),
      psd.prior = list(),
      boundary.prior = list(),
      data = transformed_data,
      X = matrix(0, nrow = 0, ncol = 0),
      B = matrix(0, nrow = 0, ncol = 0),
      P = matrix(0, nrow = 0, ncol = 0),
      initial_location = 0,
      region = range(c(0, transformed_x))
    )

    design <- as.matrix(local_poly(temp_instance))
    precision <- as.matrix(compute_weights_precision(temp_instance))
    return(design %*% solve(precision, t(design)))
  }

  stop("Unsupported model.")
}

frobenius_relative_error <- function(approximate_cov, exact_cov){
  sqrt(sum((approximate_cov - exact_cov)^2) / sum(exact_cov^2))
}

max_relative_error <- function(approximate_cov, exact_cov){
  max(abs(approximate_cov - exact_cov)) / max(abs(exact_cov))
}

variance_relative_error <- function(approximate_cov, exact_cov){
  max(abs(diag(approximate_cov) - diag(exact_cov))) / max(abs(diag(exact_cov)))
}

format_covariance_table <- function(table_df){
  table_df %>%
    mutate(
      fem_accuracy = if_else(
        model == "mgp",
        formatC(fem_accuracy, format = "fg", digits = 4),
        "not used"
      ),
      a = formatC(a, format = "fg", digits = 4),
      c_shift = formatC(c_shift, format = "fg", digits = 4),
      frobenius_rel_error = signif(frobenius_rel_error, 6),
      max_rel_error = signif(max_rel_error, 6),
      variance_rel_error = signif(variance_rel_error, 6)
    ) %>%
    rename(
      model = model,
      a = a,
      c = c_shift,
      k = k,
      `FEM accuracy` = fem_accuracy,
      `Relative Frobenius error` = frobenius_rel_error,
      `Max relative error` = max_rel_error,
      `Variance relative error` = variance_rel_error
    )
}
```

## Run the Comparison

For `tiwp2`, the column labelled `FEM accuracy` is shown as `not used` rather
than `NA`, because that FEM construction does not rely on the separate
numerical integration step used by `Compute_Prec()` for `mGP`.

```{r}
comparison_results <- parameter_grid %>%
  mutate(
    exact_covariance = pmap(
      list(model, a, c_shift),
      ~ build_exact_prior_covariance(..1, x = evaluation_grid, a = ..2, c_shift = ..3)
    )
  ) %>%
  select(-exact_covariance) %>%
  tidyr::crossing(k = k_values) %>%
  mutate(
    fem_accuracy = if_else(model == "mgp", 0.4 / k, NA_real_)
  ) %>%
  mutate(
    exact_covariance = pmap(
      list(model, a, c_shift),
      ~ build_exact_prior_covariance(..1, x = evaluation_grid, a = ..2, c_shift = ..3)
    ),
    fem_covariance = pmap(
      list(model, a, c_shift, k, fem_accuracy),
      ~ build_fem_prior_covariance(
        ..1,
        x = evaluation_grid,
        a = ..2,
        c_shift = ..3,
        k = ..4,
        mgp_accuracy = ifelse(is.na(..5), 0.002, ..5)
      )
    ),
    frobenius_rel_error = map2_dbl(fem_covariance, exact_covariance, frobenius_relative_error),
    max_rel_error = map2_dbl(fem_covariance, exact_covariance, max_relative_error),
    variance_rel_error = map2_dbl(fem_covariance, exact_covariance, variance_relative_error)
  ) %>%
  select(model, a, c_shift, k, fem_accuracy, frobenius_rel_error, max_rel_error, variance_rel_error)

knitr::kable(
  format_covariance_table(comparison_results),
  caption = "Prior covariance comparison across model, prior setting, and FEM refinement level.",
  align = c("l", "r", "r", "r", "l", "r", "r", "r")
)
```

```{r}
dir.create("output", showWarnings = FALSE)
write.csv(
  comparison_results,
  file = "output/prior_covariance_convergence_metrics.csv",
  row.names = FALSE
)
```

## Error Trends

The main quantity of interest is the relative covariance error as the FEM
resolution grows. For `mGP`, that means increasing `k` while decreasing the
integration step size at the same time.

```{r}
comparison_results %>%
  mutate(
    display_k = case_when(
      c_shift == 0.5 ~ k - 4,
      c_shift == 1.5 ~ k + 4
    )
  ) %>%
  ggplot(aes(
    x = display_k,
    y = frobenius_rel_error,
    color = factor(c_shift),
    group = c_shift,
    linetype = factor(c_shift),
    shape = factor(c_shift)
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.1) +
  facet_grid(model ~ a, scales = "free_y") +
  scale_x_continuous(breaks = k_values, labels = k_values) +
  scale_y_log10() +
  scale_linetype_manual(values = c("solid", "22")) +
  scale_shape_manual(values = c(16, 17)) +
  labs(
    x = "Number of FEM basis functions (k)",
    y = "Relative Frobenius error",
    color = "c",
    linetype = "c",
    shape = "c",
    title = "FEM prior covariance approaches the exact state-space covariance"
  ) +
  theme_bw()
```

For `tiwp2`, the two `c` curves are almost identical under the maximum
entrywise relative error metric. To keep both visible in the plot below, the
displayed `k` positions are given a very small horizontal offset by `c`, and
the two cases are also drawn with different line types. The reported numerical
values themselves are unchanged.

```{r}
comparison_results %>%
  mutate(
    display_k = case_when(
      c_shift == 0.5 ~ k - 4,
      c_shift == 1.5 ~ k + 4
    )
  ) %>%
  ggplot(aes(
    x = display_k,
    y = max_rel_error,
    color = factor(c_shift),
    group = c_shift,
    linetype = factor(c_shift),
    shape = factor(c_shift)
  )) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.1) +
  facet_grid(model ~ a, scales = "free_y") +
  scale_x_continuous(breaks = k_values, labels = k_values) +
  scale_y_log10() +
  scale_linetype_manual(values = c("solid", "22")) +
  scale_shape_manual(values = c(16, 17)) +
  labs(
    x = "Number of FEM basis functions (k)",
    y = "Maximum relative entrywise error",
    color = "c",
    linetype = "c",
    shape = "c",
    title = "Entrywise covariance errors shrink with finer FEM representations"
  ) +
  theme_bw()
```

## Largest-k Summary

```{r}
largest_k_summary <- comparison_results %>%
  group_by(model, a, c_shift) %>%
  filter(k == max(k)) %>%
  ungroup() %>%
  arrange(model, a, c_shift)

knitr::kable(
  format_covariance_table(largest_k_summary),
  caption = "Largest-refinement summary for each combination of model, a, and c.",
  align = c("l", "r", "r", "r", "l", "r", "r", "r")
)
```

## Takeaway

Across the tested combinations of `a` and `c`, the FEM covariance gets close to
the exact state-space covariance for both `mGP` and `tIWP2`.

The overall pattern is stable:

- finer FEM resolutions reduce the covariance error,
- for `mGP`, the refinement needs to include both larger `k` and a smaller
  `accuracy` step in `Compute_Prec()`,
- the variance profiles line up with the exact prior,
- and both `mGP` and `tIWP2` behave coherently under the same comparison rule.
