---
title: "Predictive Standard Deviation"
author: "Ziang Zhang"
date: "2024-06-27"
output: workflowr::wflow_html
header-includes:
  - \usepackage{amsthm}
  - \newtheorem{corollary}{Corollary}
editor_options:
  chunk_output_type: console
---

## **Introduction**

Recall that the M-GP is defined by:
\[L_af(x) = \sigma \xi(x),\]

where $L_a = D^2 - \alpha(x)D$ and $\alpha(x) = -\frac{1}{a(x+c)}$.

The monotone function implied by $\alpha(x)$ is $m = D^{-1}[\exp(D^{-1}\alpha)]$.

The t-IWP2 can be defined by a similar SDE:
\[\tilde{L}_af(x) = \sigma \xi(x),\]

where $\tilde{L}_a = m'(x)^{-3/2}L_a$.

We want to put prior on the standard deviation parameter $\sigma$ hence on the function $f$ in an interpretable way. 
One way to do this is through the notion of $h$-step predictive standard deviation (PSD) which is defined as:
\[\sigma_x(h) = \text{SD}[f(x+h)|f(u):u\leq x].\]

For process defined by time invariant differential operator, the PSD is invariant to the choice of $x$. 
For example, the h-step PSD for an IWP-2 can be shown as $\sigma\sqrt{h^3/3}$ ([Zhang et.al, 2024](https://www.tandfonline.com/doi/full/10.1080/10618600.2023.2289532)).
However for the M-GP and t-IWP2, since the differential operator involves $\alpha(x)$, the PSD is dependent on $x$.

## **PSD for M-GP and t-IWP2**

In this section, we study the behavior of PSD for M-GP and t-IWP2. 
Let $G_t(\tau)$ denotes the Green function of the operator $L_a$, the PSD $\sigma_x(h)$ can be written as:

\[\sigma^2_x(h) = \sigma^2\int_{x}^{x + h}G^2_{x+h}(t)dt.\]

Recall for the M-GP, the Green function is given by:
\[G_t(\tau) = \frac{1}{m'(\tau)}[m(t)-m(\tau)].\]

Using the initial-value Green function formula in the [textbook](https://people.uncw.edu/hermanr/mat463/ODEBook/Book/ODE2.pdf#page=280), Equation 8.38, we can show that the Green function of t-IWP2 is given by:

\[\tilde{G}_t(\tau) = m'(\tau)^{1/2}[m(t)-m(\tau)].\]

For t-IWP2, the corresponding variance is
\[
\sigma_x^2(h)
=\sigma^2\int_x^{x+h}\tilde{G}_{x+h}^2(t)\,dt
=\frac{\sigma^2}{3}\bigl[m(x+h)-m(x)\bigr]^3.
\]
The last equality follows by substituting $v=m(t)$, using $m'(t)>0$.

### *PSD of M-GP has a nice limit*

As mentioned above, the PSD of M-GP generally depends on both the location $x$ and the step size $h$.
In certain cases, we do not have a well-informed choice of the best location $x$ to put the prior on $\sigma$. 
Fortunately, we can show that the PSD of M-GP has a nice limit as $x\to \infty$, which defines a (limiting) PSD that does not require specifying the location $x$.

> **Corollary 1 (Limiting PSD of M-GP):**
>
> Fix $h>0$, $\sigma\geq0$, and a finite real $a\neq0$.
> Assume $\alpha(x)=-\frac{1}{a(x+c)}$ on $[x_0,\infty)$, where $x_0+c>0$.
> For the M-GP defined by $\alpha(x)$,
> \[\sigma^2(h)=\lim_{x\to\infty}\sigma_x^2(h)=\frac{\sigma^2h^3}{3}.\]
>

The proof of corollary 1 is given in the appendix. 
The corollary shows that although the PSD of M-GP $\sigma_x(h)$ depends on the location $x$, the limiting PSD $\sigma(h) = \lim_{x \to \infty} \sigma_x(h)$ is **exactly the same as the PSD of IWP-2**, which does not depend on the location $x$.

### *Limiting behavior for PSD of t-IWP2*

It might be tempting to think that the PSD of t-IWP2 also converges to the PSD of IWP2 as $x\to \infty$. 
However, the following corollary shows that this is not the case.

> **Corollary 2 (Limiting PSD of t-IWP2):**
>
> Fix $h>0$, $\sigma\geq0$, and a finite real $a>0$.
> Assume $\alpha(x)=-\frac{1}{a(x+c)}$ on $[x_0,\infty)$, where $x_0+c>0$.
> For t-IWP2 defined by a fixed increasing transformation $m$ with
> $m'(x)>0$ and $m''(x)/m'(x)=\alpha(x)$,
> \[\sigma^2(h)=\lim_{x\to\infty}\sigma_x^2(h)=0.\]
>

The proof of corollary 2 is also given in the appendix.

Conditional on the current function value and slope, t-IWP2 allows progressively less deviation from the implied base-model continuation over a fixed interval of length $h$, with the predictive variance tending to zero as $x\to\infty$.

## **PSD Computation**

In this section, we will show how to compute the PSD for M-GP.

The function `PSD_compute` provides a unified PSD helper across supported smooth models. For M-GP, use `model = "mgp"` and supply both the evaluation location $x$ and step size $h$.

Assume we know the value of $\sigma$, then $\sigma_x(h)$ can be computed as:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(Matrix)
library(knitr)
devtools::load_all("BayesGP", quiet = TRUE)
```

```{r}
sigma <- 5
sigma * PSD_compute(model = "mgp", x = 0, h = 2, c = 1, a = 2)
```

Now, suppose we have an Exponential prior on $\sigma_x(h)$ parametrized as:
\[
\mathbb{P}(\sigma_x(h) > u) = \alpha.
\]
We can use the function `prior_conversion_mgp` to convert the prior to an Exponential prior on $\sigma$.

For example, assume we have a prior $\mathbb{P}(\sigma_0(1) > 5) = 0.1$, then the corresponding prior on $\sigma$ is computed as:
```{r}
prior_psd <- list(u = 5, alpha = 0.1)
prior_conversion_mgp(prior_psd, x = 0, h = 1, c = 1, a = 2)
```

In this case, it corresponds to an Exponential prior $\mathbb{P}(\sigma > 9.828196) = 0.1$.

The same PSD-scale target can also be supplied directly inside a near-monotone
term specification, for example:

```{r eval = FALSE}
model_fit(
  y ~ f(
    x,
    model = "mgp",
    computation = "state-space",
    a = 2,
    c = 1,
    sd.prior = list(
      prior = "exp",
      param = list(u = 5, alpha = 0.1),
      h = 1,
      x = 0
    )
  ),
  data = data_train
)
```

## **Appendix**

At an interior starting location, the past trajectory determines the current function value and slope. Thus the PSD variance can be computed by conditioning on the state $(f(x),f'(x))$.

### *Proof of Corollary 1*

Since $m''/m'=\alpha$,
\[
m'(x)=K(x+c)^{-1/a},\qquad K>0.
\]
For $0\leq\tau\leq h$, the Green function therefore satisfies
\[
\begin{aligned}
G_{x+h}(x+\tau)
&=\frac{m(x+h)-m(x+\tau)}{m'(x+\tau)}\\
&=\int_\tau^h\frac{m'(x+s)}{m'(x+\tau)}\,ds\\
&=\int_\tau^h\left(\frac{x+s+c}{x+\tau+c}\right)^{-1/a}\,ds.
\end{aligned}
\]
On the triangle $0\leq\tau\leq s\leq h$,
\[
\sup_{0\leq\tau\leq s\leq h}
\left|\frac{x+s+c}{x+\tau+c}-1\right|
\leq\frac{h}{x+c}\longrightarrow0.
\]
For any fixed finite $a\neq0$, continuity of $z\mapsto z^{-1/a}$ at $z=1$
implies that the integrand converges uniformly to 1. Hence
$G_{x+h}(x+\tau)\to h-\tau$ uniformly on $[0,h]$.
The same holds for the squared Green function, so
\[
\begin{aligned}
\lim_{x\to\infty}\sigma_x^2(h)
&=\lim_{x\to\infty}\operatorname{Var}\bigl[f(x+h)\mid f(x),f'(x)\bigr]\\
&=\lim_{x\to\infty}\sigma^2\int_0^h G_{x+h}^2(x+\tau)\,d\tau\\
&=\sigma^2\int_0^h(h-\tau)^2\,d\tau
=\frac{\sigma^2h^3}{3}.
\end{aligned}
\]
This proves Corollary 1. $\square$

### *Proof of Corollary 2*

Write $f(x)=\tilde f(m(x))$, where $D^2\tilde f(s)=\sigma\xi(s)$.
Since
\[
f'(x)=\tilde f'(m(x))m'(x),\qquad m'(x)>0,
\]
conditioning on $(f(x),f'(x))$ is equivalent to conditioning on the IWP2
state $(\tilde f(m(x)),\tilde f'(m(x)))$.
The transformed prediction interval has length $\Delta_x=m(x+h)-m(x)$.
Using the IWP2 predictive variance gives
\[
\sigma_x^2(h)
=\operatorname{Var}\bigl[f(x+h)\mid f(x),f'(x)\bigr]
=\frac{\sigma^2}{3}[m(x+h)-m(x)]^3.
\]
Moreover, $m''/m'=-1/[a(x+c)]$ implies
\[
m'(x)=K(x+c)^{-1/a},\qquad K>0.
\]
For $0<a<\infty$, $m'(x)\to0$. By the mean value theorem, some
$\theta_x\in(0,1)$ satisfies
\[
m(x+h)-m(x)=h\,m'(x+\theta_x h)\longrightarrow0,
\]
since $x+\theta_x h\to\infty$. Consequently,
\[
\lim_{x\to\infty}\sigma_x^2(h)
=\lim_{x\to\infty}\frac{\sigma^2}{3}[m(x+h)-m(x)]^3=0.
\]
This proves Corollary 2. $\square$
