---
title: "Predictive Standard Deviation"
author: "Ziang Zhang"
date: "2024-06-27"
output: workflowr::wflow_html
header-includes:
  - \usepackage{amsthm}
  - \newtheorem{corollary}{Corollary}
editor_options:
  chunk_output_type: console
---

<!-- PSD tracked-revision display: start -->
<style>
.psd-change-delete, .psd-change-add {
  padding: 0.6em 0.8em;
  margin: 0.6em 0;
  border-left: 3px solid;
}
.psd-change-delete {
  color: #8a3341;
  background: #fff1f2;
  border-color: #b94a48;
  text-decoration: line-through;
}
.psd-change-delete p, .psd-change-delete .math {
  text-decoration: line-through;
}
.psd-change-delete .math.display {
  display: block;
  position: relative;
}
.psd-change-delete .math.display::after {
  content: "";
  position: absolute;
  left: 0;
  right: 0;
  top: 50%;
  border-top: 1px solid #b94a48;
  pointer-events: none;
}
.psd-change-add {
  background: #fff3a3;
  border-color: #b8860b;
}
</style>

::: {#psd-review-legend .psd-change-add}
**Tracked revision (2026-10-04):** Struck-through text records removed original
wording or formulas. Yellow highlights mark rewritten or added content.
Unmarked content is retained from the original page.
:::
<!-- PSD tracked-revision display: end -->

## **Introduction**

Recall that the M-GP is defined by:
\[L_af(x) = \sigma \xi(x),\]

::: {#operator-curvature-delete .psd-change-delete}

where $L_a = D^2 - \alpha(x)D$ and $\alpha(x) = -1/a(x+c)$. 

:::

::: {#operator-curvature-add .psd-change-add}

where $L_a = D^2 - \alpha(x)D$ and $\alpha(x) = -\frac{1}{a(x+c)}$.

:::

The monotone function implied by $\alpha(x)$ is $g = D^{-1}[\exp(D^{-1}\alpha)]$.

The t-IWP2 can be defined by a similar SDE:
\[\tilde{L}_af(x) = \sigma \xi(x),\]

::: {#operator-name-delete .psd-change-delete}

where $\tilde{L}_a = g'(x)^{-3/2}L_\alpha$.

:::

::: {#operator-name-add .psd-change-add}

where $\tilde{L}_a = g'(x)^{-3/2}L_a$.

:::


We want to put prior on the standard deviation parameter $\sigma$ hence on the function $f$ in an interpretable way. 
One way to do this is through the notion of $h$-step predictive standard deviation (PSD) which is defined as:
\[\sigma_x(h) = \text{SD}[f(x+h)|f(u):u\leq x].\]

::: {#psd-conditions-add .psd-change-add}

We consider a right-unbounded domain $[x_0,\infty)$ with $x_0+c>0$,
and fix $\sigma\geq0$, $h>0$, and an increasing transformation $g$ with
$g'(x)>0$. We also write $m=g$ below. The transformation's scale and
normalization are held fixed when taking the limit in $x$.
For interior starting locations, the exact past trajectory determines the
current function value and derivative. At the initial boundary, the innovation
PSD also conditions on the initial derivative (or uses a fixed initial state).
It measures uncertainty from future
innovations, rather than the full posterior uncertainty given noisy data.

:::

For process defined by time invariant differential operator, the PSD is invariant to the choice of $x$. 
For example, the h-step PSD for an IWP-2 can be shown as $\sigma\sqrt{h^3/3}$ ([Zhang et.al, 2024](https://www.tandfonline.com/doi/full/10.1080/10618600.2023.2289532)).
However for the M-GP and t-IWP2, since the differential operator involves $\alpha(x)$, the PSD is dependent on $x$.

## **PSD for M-GP and t-IWP2**

In this section, we study the behavior of PSD for M-GP and t-IWP2. 
Let $G_t(\tau)$ denotes the Green function of the operator $L_a$, the PSD $\sigma_x(h)$ can be written as:

::: {#variance-scale-delete .psd-change-delete}

\[\sigma^2_x(h) = \int_{x}^{x + h}G^2_{x+h}(t)dt.\]

:::

::: {#variance-scale-add .psd-change-add}

\[\sigma^2_x(h) = \sigma^2\int_{x}^{x + h}G^2_{x+h}(t)dt.\]

:::

Recall for the M-GP, the Green function is given by:
\[G_t(\tau) = \frac{1}{g'(\tau)}[g(t)-g(\tau)].\]

::: {#green-reference-delete .psd-change-delete}

Using property of the Green function (see [textbook](https://people.uncw.edu/hermanr/mat463/odebook/) Equation 8.36), we can show that the Green function of t-IWP2 is given by:

:::

::: {#green-reference-add .psd-change-add}

Using the initial-value Green function formula in the [textbook](https://people.uncw.edu/hermanr/mat463/ODEBook/Book/ODE2.pdf#page=280), Equation 8.38, we can show that the Green function of t-IWP2 is given by:

:::

\[\tilde{G}_t(\tau) = g'(\tau)^{1/2}[g(t)-g(\tau)].\]

::: {#tiwp2-exact-psd-add .psd-change-add}

For t-IWP2, the corresponding variance is
\[
\sigma_x^2(h)
=\sigma^2\int_x^{x+h}\tilde{G}_{x+h}^2(t)\,dt
=\frac{\sigma^2}{3}\bigl[m(x+h)-m(x)\bigr]^3.
\]
The last equality follows by substituting $v=m(t)$, using $m'(t)>0$.

:::


### *PSD of M-GP has a nice limit*

As mentioned above, the PSD of M-GP generally depends on both the location $x$ and the step size $h$.
In certain cases, we do not have a well-informed choice of the best location $x$ to put the prior on $\sigma$. 
Fortunately, we can show that the PSD of M-GP has a nice limit as $x\to \infty$, which defines a (limiting) PSD that does not require specifying the location $x$.

> **Corollary 1 (Limiting PSD of M-GP):**
>

> ::: {#mgp-limit-conditions-delete .psd-change-delete}
>
> Assume $\alpha(x) = -\frac{1}{a(x+c)}$ where $a \neq 0$. For the M-GP defined by $\alpha(x)$ \[\sigma^2(h) = \lim_{x \to \infty}\sigma_x^2(h) = \frac{\sigma^2 h^3}{3}.\]
>
> :::

> ::: {#mgp-limit-conditions-add .psd-change-add}
>
> Fix $h>0$, $\sigma\geq0$, and a finite real $a\neq0$.
> Assume $\alpha(x)=-\frac{1}{a(x+c)}$ on the domain specified above.
> For the M-GP defined by $\alpha(x)$,
> \[\sigma^2(h)=\lim_{x\to\infty}\sigma_x^2(h)=\frac{\sigma^2h^3}{3}.\]
>
> :::


The proof of corollary 1 is given in the appendix. 
The corollary shows that although the PSD of M-GP $\sigma_x(h)$ depends on the location $x$, the limiting PSD $\sigma(h) = \lim_{x \to \infty} \sigma_x(h)$ is **exactly the same as the PSD of IWP-2**, which does not depend on the location $x$.

### *Limiting behavior for PSD of t-IWP2*

It might be tempting to think that the PSD of t-IWP2 also converges to the PSD of IWP2 as $x\to \infty$. 
However, the following corollary shows that this is not the case.

> **Corollary 2 (Limiting PSD of t-IWP2):**
>

> ::: {#tiwp2-limit-conditions-delete .psd-change-delete}
>
> Assume $\alpha(x) = -\frac{1}{a(x+c)}$ where $a \geq 0$. For the t-IWP2 defined by a monotone transformation $m(x)$ such that $m''(x)/m'(x) = \alpha(x)$, the limiting PSD is \[\sigma^2(h) = \lim_{x \to \infty}\sigma_x^2(h) = 0.\]
>
> :::

> ::: {#tiwp2-limit-conditions-add .psd-change-add}
>
> Fix $h>0$, $\sigma\geq0$, and a finite real $a>0$.
> Assume $\alpha(x)=-\frac{1}{a(x+c)}$ on the domain specified above.
> For t-IWP2 defined by a fixed increasing transformation $m$ with
> $m'(x)>0$ and $m''(x)/m'(x)=\alpha(x)$,
> \[\sigma^2(h)=\lim_{x\to\infty}\sigma_x^2(h)=0.\]
>
> :::


The proof of corollary 2 is also given in the appendix.

::: {#local-versus-global-delete .psd-change-delete}

This corollary implies as the starting time increases, the PSD of t-IWP2 converges to zero, which means there is no uncertainty in the functional form of the $f$ as $x\to \infty$.

In other words, the seemingly naive approach of transforming the coordinate of an IWP2 through a monotone function actually puts a **very strong prior** on the unknown function $f$.
Basically it assumes there will be (almost) **no deviation from the base model** $\text{span}\{1,m(x)\}$ at the right end of the domain when $x$ is large.

:::

::: {#local-versus-global-add .psd-change-add}

For every fixed $h>0$, this corollary shows that the conditional
uncertainty from innovations over $[x,x+h]$ decreases to zero as the
starting location $x$ increases. Relative to M-GP at fixed process SD,
t-IWP2 therefore has vanishing local innovation variability on intervals
of a fixed length on the original scale.

This conclusion does not imply that the marginal uncertainty of $f(x)$,
its posterior uncertainty given noisy data, or its accumulated deviation
from a fixed base function vanishes. Randomness accumulated before $x$
and uncertainty in the current state can still contribute to these quantities.
It also does not establish a limit for extrapolation from a fixed observation
endpoint, where the prediction horizon increases with the target location.

For example, take $a=2$, $c=1$, $\sigma>0$, and
$m(x)=2(\sqrt{x+1}-1)$, with $m(0)=0$ and $m'(0)=1$.
With a fixed initial function value and derivative, write
$f(x)=\beta_0+\beta_1m(x)+r(x)$, where $r$ is an IWP-2 on the transformed
scale with zero initial value and derivative. Then
\[
\operatorname{Var}[f(x)\mid\beta_0,\beta_1]
=\operatorname{Var}[r(x)]=\frac{\sigma^2m(x)^3}{3}\longrightarrow\infty,
\]
while
\[
\sigma_x^2(h)=\frac{\sigma^2[m(x+h)-m(x)]^3}{3}\longrightarrow0.
\]
Thus, shrinking local innovations can coexist with increasing accumulated
uncertainty and deviation from the boundary-implied base function.

:::




## **PSD Computation**

In this section, we will show how to compute the PSD for M-GP.

The function `PSD_compute` provides a unified PSD helper across supported smooth models. For M-GP, use `model = "mgp"` and supply both the evaluation location $x$ and step size $h$.

Assume we know the value of $\sigma$, then $\sigma_x(h)$ can be computed as:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(Matrix)
library(knitr)
devtools::load_all("BayesGP", quiet = TRUE)
```

```{r}
sigma <- 5
sigma * PSD_compute(model = "mgp", x = 0, h = 2, c = 1, a = 2)
```

Now, suppose we have an Exponential prior on $\sigma_x(h)$ parametrized as:
\[
\mathbb{P}(\sigma_x(h) > u) = \alpha.
\]
We can use the function `prior_conversion_mgp` to convert the prior to an Exponential prior on $\sigma$.

For example, assume we have a prior $\mathbb{P}(\sigma_0(1) > 5) = 0.1$, then the corresponding prior on $\sigma$ is computed as:
```{r}
prior_psd <- list(u = 5, alpha = 0.1)
prior_conversion_mgp(prior_psd, x = 0, h = 1, c = 1, a = 2)
```

In this case, it corresponds to an Exponential prior $\mathbb{P}(\sigma > 9.828196) = 0.1$.

The same PSD-scale target can also be supplied directly inside a near-monotone
term specification, for example:

```{r eval = FALSE}
model_fit(
  y ~ f(
    x,
    model = "mgp",
    computation = "state-space",
    a = 2,
    c = 1,
    sd.prior = list(
      prior = "exp",
      param = list(u = 5, alpha = 0.1),
      h = 1,
      x = 0
    )
  ),
  data = data_train
)
```

## **Appendix**

### *Proof of Corollary 1*

To prove the corollary, we will utilize the following two lemmas.


> ::: {#lemma-one-domain-delete .psd-change-delete}
>
> **Lemma 1:** Assume $h\geq\tau>0$, then for any $x$:
>
> :::

> ::: {#lemma-one-domain-add .psd-change-add}
>
> **Lemma 1:** Under the assumptions of Corollary 1, for fixed $h>0$ and $0\leq\tau\leq h$:
>
> :::

> \[\lim_{x\to \infty} \frac{g(x+h)-g(x+\tau)}{g'(x+\tau)} = h-\tau.\]

*Proof of Lemma 1:*

::: {#lemma-one-endpoint-add .psd-change-add}

When $\tau=h$, the numerator is zero and the claim is immediate. We now consider $\tau<h$.

:::

Since $g$ is a continuously differentiable, by mean value theorem we know 
\[g(x+h) - g(x+\tau)=g'(x+\nu) (h-\tau),\]

::: {#mean-value-point-delete .psd-change-delete}

where $\nu \in (x+\tau, x+h)$.

:::

::: {#mean-value-point-add .psd-change-add}

where $\nu \in (\tau,h)$.

:::



::: {#lemma-one-both-signs-delete .psd-change-delete}

Assume without the loss of generality that $a >0$ so $g$ is concave function that is monotonically increasing. 
Then we have \[g'(x+h) (h-\tau) \leq g'(x+\nu) (h-\tau) \leq g'(x+\tau) (h-\tau).\]

Since by definition $h = g'$, it is clear for the right hand side we have \[\lim_{x\to \infty} \frac{g'(x+\tau) (h-\tau)}{g'(x+\tau)} = h-\tau.\]

:::

::: {#lemma-one-both-signs-add .psd-change-add}

Both signs of $a$ can be handled by monotonicity of $g'$.
For $a>0$, $g'$ is decreasing, so
\[
\frac{g'(x+h)}{g'(x+\tau)}
\leq\frac{g'(x+\nu)}{g'(x+\tau)}\leq1.
\]
For $a<0$, $g'$ is increasing, so
\[
1\leq\frac{g'(x+\nu)}{g'(x+\tau)}
\leq\frac{g'(x+h)}{g'(x+\tau)}.
\]
In either case, the mean-value ratio is bounded by 1 and the endpoint ratio.

:::



::: {#lemma-one-ratio-delete .psd-change-delete}

For the left hand side, note that:

:::

::: {#lemma-one-ratio-add .psd-change-add}

For either sign of $a$, the endpoint ratio satisfies:

:::


\begin{equation}
\begin{aligned}
\lim_{x \to \infty}\frac{g'(x+h)}{g'(x+\tau)} &= \lim_{x \to \infty}\exp \bigg(\int_{x+\tau}^{x+h} \alpha(t) dt\bigg) \\
&= \lim_{x \to \infty}\exp \bigg(\int_{x+\tau}^{x+h} \frac{-1}{a(t+c)} dt\bigg) \\
&= \lim_{x \to \infty}\exp \bigg(\frac{1}{a} \ln \bigg(\frac{x+\tau+c}{x+h+c}\bigg)\bigg) \\
&=  \lim_{x \to \infty} \bigg(\frac{x+\tau+c}{x+h+c}\bigg)^{1/a} = 1.
\end{aligned}
\end{equation}

So by squeezing theorem, we have \[\lim_{x\to \infty} \frac{g(x+h)-g(x+\tau)}{g'(x+\tau)} = h-\tau.\] 
This lemma is hence proved. \qed

> **Lemma 2:** Assume $h>0$, then for any $x$:
> \[\lim_{x\to \infty} \int_{x}^{x+h} G^2_{x+h}(t)dt = \int_{0}^{h} \lim_{x\to \infty} G^2_{x+h}(x + t)dt.\]

*Proof of Lemma 2:*
The proof of this lemma is a simple application of the dominated convergence theorem.

First it is clear through a simple change of variable that \[\int_{x}^{x+h} G^2_{x+h}(t)dt = \int_{0}^{h} G^2_{x+h}(x + t)dt.\]


::: {#lemma-two-dominator-delete .psd-change-delete}

Again, assume without the loss of generality that $a >0$ so $g$ is concave.
We have
\begin{equation}
\begin{aligned}
\left|G_{x+h}(x + t)\right| &= \left|\frac{g(x+h) - g(x + t)}{g'(x+t)}\right| \\
&\leq \left|\frac{g'(x+t)(h-t)}{g'(x+t)}\right| \\
&= |h - t|.
\end{aligned}
\end{equation}

Since $|h - t|$ is integrable over $[0,h]$, this lemma follows from the dominated convergence theorem.

:::

::: {#lemma-two-dominator-add .psd-change-add}

For either sign of $a\neq0$, we have
\[
G_{x+h}(x+t)=\int_t^h\frac{g'(x+s)}{g'(x+t)}\,ds,
\qquad
\frac{g'(x+s)}{g'(x+t)}
=\left(\frac{x+s+c}{x+t+c}\right)^{-1/a}.
\]
Choose a fixed $X\geq x_0$ with $X+c\geq h$. For $x\geq X$ and
$0\leq t\leq s\leq h$, the ratio inside the power lies in $[1,2]$.
Consequently, with $K_a=2^{|1/a|}$,
\[
0\leq G_{x+h}(x+t)\leq K_a(h-t),
\qquad
G_{x+h}^2(x+t)\leq K_a^2(h-t)^2.
\]
This squared bound is independent of $x$ and integrable over $[0,h]$.
The dominated convergence theorem therefore applies for both positive and
negative $a$.

:::


With the above two lemmas, we can show that:
\begin{equation}
\begin{aligned}
\lim_{x\to \infty} \int_{x}^{x+h} G^2_{x+h}(t)dt &= \int_{0}^{h} \lim_{x\to \infty} G^2_{x+h}(x + t)dt \\
&= \int_{0}^{h} \bigg\{\lim_{x\to \infty} \frac{1}{g'(x+t)}[g(x+h)-g(x+t)]\bigg\}^2dt \\
&= \int_{0}^{h} (h-t)^2dt \\
&= \frac{h^3}{3}.
\end{aligned}
\end{equation}

The proof of corollary 1 is hence completed.

::: {#mgp-proof-scale-add .psd-change-add}

The integral above is the unit-noise variance factor. Multiplying its limit by $\sigma^2$ gives the PSD variance asserted in Corollary 1.

:::



### *Proof of Corollary 2*

To prove this corollary, we first note that the Green function of t-IWP2 is given by:
\[\tilde{G}_t(\tau) = g'(\tau)^{1/2}[g(t)-g(\tau)].\]


::: {#tiwp2-proof-delete .psd-change-delete}

When $a > 0$, we know $|g(x+h) - g(x+\tau)| \leq g'(x+\tau)(h-\tau)$ if $h\geq \tau$.
So we have:
\begin{equation}
\begin{aligned}
g'(x + \tau)^{1/2}[g(x + h)-g(x + \tau)] &\leq g'(x + \tau)^{3/2}(h-\tau)\\
&= (h-\tau) \exp \bigg[ - \int_{0}^{x+\tau} \frac{1}{a(t+c)}dt \bigg]^{3/2} \\
&= (h-\tau) \exp \bigg[ - \frac{1}{a} \ln \bigg(\frac{x+\tau+c}{c}\bigg) \bigg]^{3/2} \\
&= (h-\tau) \bigg(\frac{c}{x+\tau+c}\bigg)^{\frac{3}{2a}}. \\
\end{aligned}
\end{equation}

Therefore, we know \[0\leq\lim_{x\to\infty} \tilde{G}^2_{x+h}(x+\tau) \leq (h-\tau)^2\lim_{x\to\infty}  \bigg(\frac{c}{x+\tau+c}\bigg)^{3/a} = 0.\]

Finally, since $\tilde{G}^2_{x+h}(x+\tau) \leq (h-\tau)^2$ which is integrable over $[0,h]$, we can apply the dominated convergence theorem.
Therefore \[\lim_{x\to\infty} \int_{x}^{x+h}\tilde{G}^2_{x+h}(t)dt = \lim_{x\to\infty}\int_{0}^{h}\tilde{G}^2_{x+h}(x + \tau)d\tau = \int_{0}^{h} \left[\lim_{x\to\infty} \tilde{G}^2_{x+h}(x+\tau) \right]d\tau = 0. \]

:::

::: {#tiwp2-proof-add .psd-change-add}

Writing $m=g$, the Green-function variance and the substitution $v=m(t)$ give
\begin{equation}
\begin{aligned}
\sigma_x^2(h)
&=\sigma^2\int_x^{x+h}m'(t)[m(x+h)-m(t)]^2\,dt\\
&=\sigma^2\int_{m(x)}^{m(x+h)}[m(x+h)-v]^2\,dv\\
&=\frac{\sigma^2}{3}[m(x+h)-m(x)]^3.
\end{aligned}
\end{equation}
Since $m''/m'=-1/[a(x+c)]$, integration gives
\[
m'(x)=K(x+c)^{-1/a},\qquad K>0,
\]
where $K$ is fixed by the chosen transformation normalization.
For finite $a>0$, $m'(x)\to0$. By the mean value theorem, for each fixed
$h>0$ there is $\theta_x\in(0,h)$ such that
\[
0<m(x+h)-m(x)=h\,m'(x+\theta_x)
\leq h\,m'(x)\longrightarrow0.
\]
Substituting this into the exact PSD formula proves the stated zero limit,
without requiring a particular value of $m'(0)$ or assuming that $0$ lies
in the domain.

:::


This completes the proof of corollary 2. 
