---
title: "Simulation 1"
author: "Ziang Zhang"
date: "2024-06-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(Matrix)
library(knitr)
devtools::load_all("BayesGP", quiet = TRUE)
c <- 1
B <- 1000
n <- 200
true_psd <- 2
sd_noise <- 2
set.seed(123)
```

## **Introduction**

In this simulation study, we study the robustness of the tIWP2 and the mGP priors.


## **Simulation A: True model is mGP**

The simulation A is designed as follows, for each of the $B = 1000$ replications:

- $f$ is simulated from a mGP process with $a = 2$, with a true PSD $\sigma_0(5) = 2$.
- For $x_i$ equally spaced between $[0,20]$, $n = 200$ observations are simulated from $y_i = f(x_i) + \epsilon_i$ with $\epsilon_i \sim N(0, 2)$. 
- The posterior $\pi(f|\boldsymbol{y})$ is obtained using both the tIWP2 and the mGP priors, only using data with $x \in [0,10]$. The prior for the smoothing parameter is chosen as an Exponential prior centered at the true PSD $\sigma_0(5)$.
- The pointwise coverage probabilities of the 50\%, 80\%, and 95\% credible interval for $f$, as well as the mean squared error and the mean relative error of the posterior mean of $f$ are all recorded for each replication.

```{r, echo=FALSE}
sim_data_once <- function(n = 199, sd_noise = 2, psd_fun = 2){
  x <- seq(0, 20, length.out = n + 1)
  sd_fun <- psd_fun / PSD_compute(model = "mgp", h = 5, x = 0, a = 2, c = c)
  f_vec <- mGP_sim(t = x, alpha = 2, c = c, initial_vec = c(0, 5), sd = sd_fun)[-1, 2]
  data.frame(x = x[-1], y = f_vec + rnorm(n, 0, sd_noise), f = f_vec)
}

evaluate_model_once <- function(data_train, data_sim, model_fit, M = 3000){
  evaluate_train_prediction_draws(
    samples = as.matrix(
      predict(model_fit, newdata = data_sim, variable = "x", only.samples = TRUE)[, -1, drop = FALSE]
    ),
    truth = data_sim$f,
    n_train = nrow(data_train)
  )
}

fit_nearmonotone_model <- function(data_train, data_sim, model_name, prior_u,
                                   computation_method = "state-space",
                                   normalized_boundary = TRUE){
  model_fit(
    formula = y ~ f(
      x,
      model = model_name,
      computation = computation_method,
      a = 2,
      c = c,
      grid = data_sim$x,
      normalized_boundary = normalized_boundary,
      sd.prior = list(prior = "exp", param = list(u = prior_u, alpha = 0.5))
    ),
    data = data_train,
    family = "gaussian",
    control.family = list(sd = sd_noise),
    M = 3000
  )
}

replicate_comparison_B_times <- function(B, n, sd_noise = 2, psd_fun = 2,
                                         observed_max = 10, parallel = FALSE,
                                         u_mgp, u_tiwp2) {
  empty_metrics <- data.frame(
    coverage_50 = numeric(0),
    coverage_80 = numeric(0),
    coverage_95 = numeric(0),
    rmse = numeric(0),
    rmle = numeric(0)
  )

  run_one_iteration <- function(){
    data_sim <- sim_data_once(n = n, sd_noise = sd_noise, psd_fun = psd_fun)
    data_train <- data_sim %>% filter(x <= observed_max)

    mGP_fit <- fit_nearmonotone_model(
      data_train = data_train,
      data_sim = data_sim,
      model_name = "mgp",
      prior_u = u_mgp,
      computation_method = "state-space",
      normalized_boundary = TRUE
    )
    tiwp2_fit <- fit_nearmonotone_model(
      data_train = data_train,
      data_sim = data_sim,
      model_name = "tiwp2",
      prior_u = u_tiwp2,
      computation_method = "state-space"
    )

    mGP_result <- evaluate_model_once(data_train = data_train, data_sim = data_sim, model_fit = mGP_fit)
    tIWP2_result <- evaluate_model_once(data_train = data_train, data_sim = data_sim, model_fit = tiwp2_fit)

    list(
      mGP_inter = mGP_result$data_inter,
      mGP_pred = mGP_result$data_pred,
      tIWP2_inter = tIWP2_result$data_inter,
      tIWP2_pred = tIWP2_result$data_pred
    )
  }

  bind_metric_rows <- function(results, field){
    valid_results <- Filter(Negate(is.null), results)
    if(length(valid_results) == 0){
      return(empty_metrics)
    }
    bind_rows(lapply(valid_results, `[[`, field))
  }

  if(!parallel) {
    results <- vector("list", B)
    for(i in seq_len(B)) {
      results[[i]] <- tryCatch(
        run_one_iteration(),
        error = function(e) {
          cat("Error at iteration", i, ":", e$message, "\n")
          NULL
        }
      )
      if(i %% 10 == 0){
        cat("Completed iteration", i, "out of", B, "\n")
      }
    }
  } else {
    library(parallel)
    cl <- makeCluster(detectCores() - 12)
    on.exit(stopCluster(cl))

    clusterExport(
      cl,
      varlist = c("sim_data_once", "evaluate_model_once", "run_one_iteration",
                  "n", "c", "psd_factor_mgp", "psd_factor_tiwp2", "true_psd",
                  "sd_noise", "observed_max", "u_mgp", "u_tiwp2", "psd_fun",
                  "fit_nearmonotone_model"),
      envir = environment()
    )

    clusterEvalQ(cl, {
      library(tidyverse)
      devtools::load_all("BayesGP", quiet = TRUE)
      TRUE
    })

    results <- parLapply(cl, seq_len(B), function(i){
      tryCatch(
        run_one_iteration(),
        error = function(e) {
          cat("Error in iteration", i, ":", e$message, "\n")
          NULL
        }
      )
    })
  }

  list(
    mGP_inter = bind_metric_rows(results, "mGP_inter"),
    mGP_pred = bind_metric_rows(results, "mGP_pred"),
    tIWP2_inter = bind_metric_rows(results, "tIWP2_inter"),
    tIWP2_pred = bind_metric_rows(results, "tIWP2_pred")
  )
}
```

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
psd_factor_tiwp2 <- true_psd / PSD_compute(model = "tiwp2", h = 5, x = 0, a = 2, c = 1)
psd_factor_mgp <- true_psd / PSD_compute(model = "mgp", h = 5, x = 0, a = 2, c = c)
all_result <- replicate_comparison_B_times(B = B, n = n, parallel = T, u_mgp = psd_factor_mgp, u_tiwp2 = psd_factor_tiwp2, sd_noise = 2, psd_fun = true_psd)
save(all_result, file = "output/sim1_all_resultA.rda")
```

```{r, echo=FALSE}
load("output/sim1_all_resultA.rda")
```

Visualize the RMSE result for interpolation and prediction:
```{r, echo=FALSE}
rmse_results <- bind_rows(
  all_result$mGP_inter %>%
    mutate(model = "mGP", type = "inter"),
  all_result$tIWP2_inter %>%
    mutate(model = "tIWP2", type = "inter"),
  all_result$mGP_pred %>%
    mutate(model = "mGP", type = "pred"),
  all_result$tIWP2_pred %>%
    mutate(model = "tIWP2", type = "pred")
) %>%
  select(rmse, model, type) %>%
  pivot_longer(cols = starts_with("rmse"),
               names_to = "rmse_level",
               values_to = "rmse_value") %>%
  mutate(rmse_level = gsub("rmse_", "", rmse_level))

rmse_results %>% ggplot(aes(x = type, y = (rmse_value), fill = model)) +
  geom_boxplot(outliers = F) +
  scale_fill_manual(values = c("mGP" = "skyblue", "tIWP2" = "salmon")) +
  labs(title = "Comparison of RMSE",
       x = "Setting",
       y = "RMSE",
       fill = "Model") +
  scale_y_log10() +
  theme_minimal()
```

Visualize the MLE result for interpolation and prediction:
```{r, echo=FALSE}
rmle_results <- bind_rows(
  all_result$mGP_inter %>%
    mutate(model = "mGP", type = "inter"),
  all_result$tIWP2_inter %>%
    mutate(model = "tIWP2", type = "inter"),
  all_result$mGP_pred %>%
    mutate(model = "mGP", type = "pred"),
  all_result$tIWP2_pred %>%
    mutate(model = "tIWP2", type = "pred")
) %>%
  select(rmle, model, type) %>%
  pivot_longer(cols = starts_with("rmle"),
               names_to = "rmle_level",
               values_to = "rmle_value") %>%
  mutate(rmle_level = gsub("rmle_", "", rmle_level))

rmle_results %>% ggplot(aes(x = type, y = rmle_value, fill = model)) +
  geom_boxplot(outliers = F) +
  scale_fill_manual(values = c("mGP" = "skyblue", "tIWP2" = "salmon")) +
  labs(title = "Comparison of mean relative error",
       x = "Setting",
       y = "relative error",
       fill = "Model") +
  theme_minimal()

```

A table of the summary statistics:
```{r, echo=FALSE}
# Aggregating mean and standard deviation for inter results
inter_summary <- bind_rows(
  all_result$mGP_inter %>%
    summarise(
      model = "mGP",
      mean_coverage_50 = mean(coverage_50),
      sd_coverage_50 = sqrt(mean(coverage_50)*(1-mean(coverage_50))/n()),
      mean_coverage_80 = mean(coverage_80),
      sd_coverage_80 = sqrt(mean(coverage_80)*(1-mean(coverage_80))/n()),
      mean_coverage_95 = mean(coverage_95),
      sd_coverage_95 = sqrt(mean(coverage_95)*(1-mean(coverage_95))/n()),
      mean_rmse = mean(rmse),
      sd_rmse = sd(rmse),
      mean_rmle = mean(rmle),
      sd_rmle = sd(rmle)
    ),
  all_result$tIWP2_inter %>%
    summarise(
      model = "tIWP2",
      mean_coverage_50 = mean(coverage_50),
      sd_coverage_50 = sqrt(mean(coverage_50)*(1-mean(coverage_50))/n()),
      mean_coverage_80 = mean(coverage_80),
      sd_coverage_80 = sqrt(mean(coverage_80)*(1-mean(coverage_80))/n()),
      mean_coverage_95 = mean(coverage_95),
      sd_coverage_95 = sqrt(mean(coverage_95)*(1-mean(coverage_95))/n()),
      mean_rmse = mean(rmse),
      sd_rmse = sd(rmse),
      mean_rmle = mean(rmle),
      sd_rmle = sd(rmle)
    )
) %>%
  mutate(case = "Inter")

# Aggregating mean and standard deviation for pred results
pred_summary <- bind_rows(
  all_result$mGP_pred %>%
    summarise(
      model = "mGP",
      mean_coverage_50 = mean(coverage_50),
      sd_coverage_50 = sqrt(mean(coverage_50)*(1-mean(coverage_50))/n()),
      mean_coverage_80 = mean(coverage_80),
      sd_coverage_80 = sqrt(mean(coverage_80)*(1-mean(coverage_80))/n()),
      mean_coverage_95 = mean(coverage_95),
      sd_coverage_95 = sqrt(mean(coverage_95)*(1-mean(coverage_95))/n()),
      mean_rmse = mean(rmse),
      sd_rmse = sd(rmse),
      mean_rmle = mean(rmle),
      sd_rmle = sd(rmle)
    ),
  all_result$tIWP2_pred %>%
    summarise(
      model = "tIWP2",
      mean_coverage_50 = mean(coverage_50),
      sd_coverage_50 = sqrt(mean(coverage_50)*(1-mean(coverage_50))/n()),
      mean_coverage_80 = mean(coverage_80),
      sd_coverage_80 = sqrt(mean(coverage_80)*(1-mean(coverage_80))/n()),
      mean_coverage_95 = mean(coverage_95),
      sd_coverage_95 = sqrt(mean(coverage_95)*(1-mean(coverage_95))/n()),
      mean_rmse = mean(rmse),
      sd_rmse = sd(rmse),
      mean_rmle = mean(rmle),
      sd_rmle = sd(rmle)
    )
) %>%
  mutate(case = "Pred")

# Combine mean and standard deviation into a single column
format_with_sd <- function(mean_val, sd_val) {
  paste0(round(mean_val, 3), " (", round(sd_val, 3), ")")
}

# Create summary table for the "Inter" scenario
inter_summary_table <- inter_summary %>%
  mutate(
    Coverage_50 = format_with_sd(mean_coverage_50, sd_coverage_50),
    Coverage_80 = format_with_sd(mean_coverage_80, sd_coverage_80),
    Coverage_95 = format_with_sd(mean_coverage_95, sd_coverage_95),
    RMSE = format_with_sd(mean_rmse, sd_rmse),
    MLE = format_with_sd(mean_rmle, sd_rmle)
  ) %>%
  select(Model = model, `Coverage 50%` = Coverage_50, `Coverage 80%` = Coverage_80, `Coverage 95%` = Coverage_95, RMSE, MLE)

# Create summary table for the "Pred" scenario
pred_summary_table <- pred_summary %>%
  mutate(
    Coverage_50 = format_with_sd(mean_coverage_50, sd_coverage_50),
    Coverage_80 = format_with_sd(mean_coverage_80, sd_coverage_80),
    Coverage_95 = format_with_sd(mean_coverage_95, sd_coverage_95),
    RMSE = format_with_sd(mean_rmse, sd_rmse),
    MLE = format_with_sd(mean_rmle, sd_rmle)
  ) %>%
  select(Model = model, `Coverage 50%` = Coverage_50, `Coverage 80%` = Coverage_80, `Coverage 95%` = Coverage_95, RMSE, MLE)

# Display the tables using kable
inter_summary_table %>%
  kable(format = "markdown", caption = "Summary for Inter Scenario")

pred_summary_table %>%
  kable(format = "markdown", caption = "Summary for Pred Scenario")
```

Visualize the coverage probabilities for interpolation and prediction:
```{r, echo=FALSE}
# Prepare the data for the Inter scenario
inter_coverage_plot_data <- inter_summary %>%
  pivot_longer(cols = starts_with("mean_coverage"),
               names_to = "Coverage_Level",
               values_to = "Coverage") %>%
  mutate(SD = case_when(
    Coverage_Level == "mean_coverage_50" ~ sd_coverage_50,
    Coverage_Level == "mean_coverage_80" ~ sd_coverage_80,
    Coverage_Level == "mean_coverage_95" ~ sd_coverage_95
  )) %>%
  mutate(
    lower = Coverage - 3 * SD,
    upper = Coverage + 3 * SD,
    Coverage_Level = recode(Coverage_Level, 
                            "mean_coverage_50" = "50%", 
                            "mean_coverage_80" = "80%", 
                            "mean_coverage_95" = "95%")
  )

ggplot(inter_coverage_plot_data, aes(x = Coverage_Level, y = Coverage, fill = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3, shape = 21, color = "black") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(width = 0.5)) +
  geom_hline(aes(yintercept = as.numeric(sub("%", "", Coverage_Level)) / 100), linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("mGP" = "skyblue", "tIWP2" = "salmon")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0.5, 0.8, 0.95)) +
  labs(title = "Interval Plot of Coverage Rates (Inter Scenario)",
       x = "Coverage Level",
       y = "Coverage Rate",
       fill = "Model") +
  theme_minimal()


# Prepare the data for the Pred scenario
pred_coverage_plot_data <- pred_summary %>%
  pivot_longer(cols = starts_with("mean_coverage"),
               names_to = "Coverage_Level",
               values_to = "Coverage") %>%
  mutate(SD = case_when(
    Coverage_Level == "mean_coverage_50" ~ sd_coverage_50,
    Coverage_Level == "mean_coverage_80" ~ sd_coverage_80,
    Coverage_Level == "mean_coverage_95" ~ sd_coverage_95
  )) %>%
  mutate(
    lower = Coverage - 3 * SD,
    upper = Coverage + 3 * SD,
    Coverage_Level = recode(Coverage_Level, 
                            "mean_coverage_50" = "50%", 
                            "mean_coverage_80" = "80%", 
                            "mean_coverage_95" = "95%")
  )

ggplot(pred_coverage_plot_data, aes(x = Coverage_Level, y = Coverage, fill = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3, shape = 21, color = "black") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(width = 0.5)) +
  geom_hline(aes(yintercept = as.numeric(sub("%", "", Coverage_Level)) / 100), linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("mGP" = "skyblue", "tIWP2" = "salmon")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0.5, 0.8, 0.95)) +
  labs(title = "Interval Plot of Coverage Rates (Pred Scenario)",
       x = "Coverage Level",
       y = "Coverage Rate",
       fill = "Model") +
  theme_minimal()
```






## **Simulation B: True model is tIWP2**

The simulation B is designed similarly as the simulation A, except now:

- $f$ is simulated from a tIWP2 process generated by $\sqrt{x+c}$, with a true PSD $\sigma_0(5) = 2$.

```{r, echo=FALSE}
set.seed(123)
sim_data_once_B <- function(n = 199, sd_noise = 2, psd_fun = 2){
  x <- seq(0, 20, length.out = n + 1)
  sd_fun <- psd_fun / PSD_compute(model = "tiwp2", h = 5, x = 0, a = 2, c = c)
  f_vec <- simulate_tiwp_boxcox(
    x = x,
    a = 2,
    c = c,
    sd = sd_fun,
    initial_level = 0,
    initial_slope = 5,
    normalized_transform = TRUE
  )[-1]
  data.frame(x = x[-1], y = f_vec + rnorm(n, 0, sd_noise), f = f_vec)
}
true_psd <- PSD_compute(model = "tiwp2", h = 5, x = 0, sd = 1, a = 2, c = 1)
psd_factor_mgp <- true_psd / PSD_compute(model = "mgp", h = 5, x = 0, a = 2, c = c)
# data_sim <- sim_data_once_B(n)
# plot(data_sim$x, data_sim$f, type = "l", col = "blue", lwd = 2, xlab = "x", ylab = "f(x)")
# points(data_sim$x, data_sim$y, col = "black")
replicate_comparison_B_times_B <- function(B, n, sd_noise = 2, psd_fun = 2,
                                           observed_max = 10, parallel = FALSE,
                                           u_mgp, u_tiwp2) {
  empty_metrics <- data.frame(
    coverage_50 = numeric(0),
    coverage_80 = numeric(0),
    coverage_95 = numeric(0),
    rmse = numeric(0),
    rmle = numeric(0)
  )

  run_one_iteration <- function(){
    data_sim <- sim_data_once_B(n = n, sd_noise = sd_noise, psd_fun = psd_fun)
    data_train <- data_sim %>% filter(x <= observed_max)

    mGP_fit <- fit_nearmonotone_model(
      data_train = data_train,
      data_sim = data_sim,
      model_name = "mgp",
      prior_u = u_mgp,
      computation_method = "state-space",
      normalized_boundary = TRUE
    )
    tiwp2_fit <- fit_nearmonotone_model(
      data_train = data_train,
      data_sim = data_sim,
      model_name = "tiwp2",
      prior_u = u_tiwp2,
      computation_method = "state-space"
    )

    mGP_result <- evaluate_model_once(data_train = data_train, data_sim = data_sim, model_fit = mGP_fit)
    tIWP2_result <- evaluate_model_once(data_train = data_train, data_sim = data_sim, model_fit = tiwp2_fit)

    list(
      mGP_inter = mGP_result$data_inter,
      mGP_pred = mGP_result$data_pred,
      tIWP2_inter = tIWP2_result$data_inter,
      tIWP2_pred = tIWP2_result$data_pred
    )
  }

  bind_metric_rows <- function(results, field){
    valid_results <- Filter(Negate(is.null), results)
    if(length(valid_results) == 0){
      return(empty_metrics)
    }
    bind_rows(lapply(valid_results, `[[`, field))
  }

  if(!parallel) {
    results <- vector("list", B)
    for(i in seq_len(B)) {
      results[[i]] <- tryCatch(
        run_one_iteration(),
        error = function(e) {
          cat("Error at iteration", i, ":", e$message, "\n")
          NULL
        }
      )
      if(i %% 10 == 0){
        cat("Completed iteration", i, "out of", B, "\n")
      }
    }
  } else {
    library(parallel)
    cl <- makeCluster(detectCores() - 12)
    on.exit(stopCluster(cl))

    clusterExport(
      cl,
      varlist = c("sim_data_once_B", "evaluate_model_once", "run_one_iteration",
                  "n", "c", "psd_factor_mgp", "psd_factor_tiwp2", "true_psd",
                  "sd_noise", "observed_max", "u_mgp", "u_tiwp2", "psd_fun",
                  "fit_nearmonotone_model"),
      envir = environment()
    )

    clusterEvalQ(cl, {
      library(tidyverse)
      devtools::load_all("BayesGP", quiet = TRUE)
      TRUE
    })

    results <- parLapply(cl, seq_len(B), function(i){
      tryCatch(
        run_one_iteration(),
        error = function(e) {
          cat("Error in iteration", i, ":", e$message, "\n")
          NULL
        }
      )
    })
  }

  list(
    mGP_inter = bind_metric_rows(results, "mGP_inter"),
    mGP_pred = bind_metric_rows(results, "mGP_pred"),
    tIWP2_inter = bind_metric_rows(results, "tIWP2_inter"),
    tIWP2_pred = bind_metric_rows(results, "tIWP2_pred")
  )
}

```

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
psd_factor_tiwp2 <- true_psd / PSD_compute(model = "tiwp2", h = 5, x = 0, a = 2, c = 1)
psd_factor_mgp <- true_psd / PSD_compute(model = "mgp", h = 5, x = 0, a = 2, c = c)
all_result <- replicate_comparison_B_times_B(B = B, n = n, parallel = T, u_mgp = psd_factor_mgp, u_tiwp2 = psd_factor_tiwp2, sd_noise = sd_noise, psd_fun = true_psd)
save(all_result, file = "output/sim1_all_resultB.rda")
```

```{r, echo=FALSE}
load("output/sim1_all_resultB.rda")
```

Visualize the RMSE result for interpolation and prediction:
```{r, echo=FALSE}
rmse_results <- bind_rows(
  all_result$mGP_inter %>%
    mutate(model = "mGP", type = "inter"),
  all_result$tIWP2_inter %>%
    mutate(model = "tIWP2", type = "inter"),
  all_result$mGP_pred %>%
    mutate(model = "mGP", type = "pred"),
  all_result$tIWP2_pred %>%
    mutate(model = "tIWP2", type = "pred")
) %>%
  select(rmse, model, type) %>%
  pivot_longer(cols = starts_with("rmse"),
               names_to = "rmse_level",
               values_to = "rmse_value") %>%
  mutate(rmse_level = gsub("rmse_", "", rmse_level))

rmse_results %>% ggplot(aes(x = type, y = (rmse_value), fill = model)) +
  geom_boxplot(outliers = F) +
  scale_fill_manual(values = c("mGP" = "skyblue", "tIWP2" = "salmon")) +
  labs(title = "Comparison of RMSE",
       x = "Setting",
       y = "RMSE",
       fill = "Model") +
  scale_y_log10() +
  theme_minimal()
```

Visualize the MLE result for interpolation and prediction:
```{r, echo=FALSE}
rmle_results <- bind_rows(
  all_result$mGP_inter %>%
    mutate(model = "mGP", type = "inter"),
  all_result$tIWP2_inter %>%
    mutate(model = "tIWP2", type = "inter"),
  all_result$mGP_pred %>%
    mutate(model = "mGP", type = "pred"),
  all_result$tIWP2_pred %>%
    mutate(model = "tIWP2", type = "pred")
) %>%
  select(rmle, model, type) %>%
  pivot_longer(cols = starts_with("rmle"),
               names_to = "rmle_level",
               values_to = "rmle_value") %>%
  mutate(rmle_level = gsub("rmle_", "", rmle_level))

rmle_results %>% ggplot(aes(x = type, y = rmle_value, fill = model)) +
  geom_boxplot(outliers = F) +
  scale_fill_manual(values = c("mGP" = "skyblue", "tIWP2" = "salmon")) +
  labs(title = "Comparison of mean relative error",
       x = "Setting",
       y = "relative error",
       fill = "Model") +
  theme_minimal()

```

A table of the summary statistics:
```{r, echo=FALSE}
# Aggregating mean and standard deviation for inter results
inter_summary <- bind_rows(
  all_result$mGP_inter %>%
    summarise(
      model = "mGP",
      mean_coverage_50 = mean(coverage_50),
      sd_coverage_50 = sqrt(mean(coverage_50)*(1-mean(coverage_50))/n()),
      mean_coverage_80 = mean(coverage_80),
      sd_coverage_80 = sqrt(mean(coverage_80)*(1-mean(coverage_80))/n()),
      mean_coverage_95 = mean(coverage_95),
      sd_coverage_95 = sqrt(mean(coverage_95)*(1-mean(coverage_95))/n()),
      mean_rmse = mean(rmse),
      sd_rmse = sd(rmse),
      mean_rmle = mean(rmle),
      sd_rmle = sd(rmle)
    ),
  all_result$tIWP2_inter %>%
    summarise(
      model = "tIWP2",
      mean_coverage_50 = mean(coverage_50),
      sd_coverage_50 = sqrt(mean(coverage_50)*(1-mean(coverage_50))/n()),
      mean_coverage_80 = mean(coverage_80),
      sd_coverage_80 = sqrt(mean(coverage_80)*(1-mean(coverage_80))/n()),
      mean_coverage_95 = mean(coverage_95),
      sd_coverage_95 = sqrt(mean(coverage_95)*(1-mean(coverage_95))/n()),
      mean_rmse = mean(rmse),
      sd_rmse = sd(rmse),
      mean_rmle = mean(rmle),
      sd_rmle = sd(rmle)
    )
) %>%
  mutate(case = "Inter")

# Aggregating mean and standard deviation for pred results
pred_summary <- bind_rows(
  all_result$mGP_pred %>%
    summarise(
      model = "mGP",
      mean_coverage_50 = mean(coverage_50),
      sd_coverage_50 = sqrt(mean(coverage_50)*(1-mean(coverage_50))/n()),
      mean_coverage_80 = mean(coverage_80),
      sd_coverage_80 = sqrt(mean(coverage_80)*(1-mean(coverage_80))/n()),
      mean_coverage_95 = mean(coverage_95),
      sd_coverage_95 = sqrt(mean(coverage_95)*(1-mean(coverage_95))/n()),
      mean_rmse = mean(rmse),
      sd_rmse = sd(rmse),
      mean_rmle = mean(rmle),
      sd_rmle = sd(rmle)
    ),
  all_result$tIWP2_pred %>%
    summarise(
      model = "tIWP2",
      mean_coverage_50 = mean(coverage_50),
      sd_coverage_50 = sqrt(mean(coverage_50)*(1-mean(coverage_50))/n()),
      mean_coverage_80 = mean(coverage_80),
      sd_coverage_80 = sqrt(mean(coverage_80)*(1-mean(coverage_80))/n()),
      mean_coverage_95 = mean(coverage_95),
      sd_coverage_95 = sqrt(mean(coverage_95)*(1-mean(coverage_95))/n()),
      mean_rmse = mean(rmse),
      sd_rmse = sd(rmse),
      mean_rmle = mean(rmle),
      sd_rmle = sd(rmle)
    )
) %>%
  mutate(case = "Pred")

# Combine mean and standard deviation into a single column
format_with_sd <- function(mean_val, sd_val) {
  paste0(round(mean_val, 3), " (", round(sd_val, 3), ")")
}

# Create summary table for the "Inter" scenario
inter_summary_table <- inter_summary %>%
  mutate(
    Coverage_50 = format_with_sd(mean_coverage_50, sd_coverage_50),
    Coverage_80 = format_with_sd(mean_coverage_80, sd_coverage_80),
    Coverage_95 = format_with_sd(mean_coverage_95, sd_coverage_95),
    RMSE = format_with_sd(mean_rmse, sd_rmse),
    MLE = format_with_sd(mean_rmle, sd_rmle)
  ) %>%
  select(Model = model, `Coverage 50%` = Coverage_50, `Coverage 80%` = Coverage_80, `Coverage 95%` = Coverage_95, RMSE, MLE)

# Create summary table for the "Pred" scenario
pred_summary_table <- pred_summary %>%
  mutate(
    Coverage_50 = format_with_sd(mean_coverage_50, sd_coverage_50),
    Coverage_80 = format_with_sd(mean_coverage_80, sd_coverage_80),
    Coverage_95 = format_with_sd(mean_coverage_95, sd_coverage_95),
    RMSE = format_with_sd(mean_rmse, sd_rmse),
    MLE = format_with_sd(mean_rmle, sd_rmle)
  ) %>%
  select(Model = model, `Coverage 50%` = Coverage_50, `Coverage 80%` = Coverage_80, `Coverage 95%` = Coverage_95, RMSE, MLE)

# Display the tables using kable
inter_summary_table %>%
  kable(format = "markdown", caption = "Summary for Inter Scenario")

pred_summary_table %>%
  kable(format = "markdown", caption = "Summary for Pred Scenario")
```

Visualize the coverage probabilities for interpolation and prediction:
```{r, echo=FALSE}
# Prepare the data for the Inter scenario
inter_coverage_plot_data <- inter_summary %>%
  pivot_longer(cols = starts_with("mean_coverage"),
               names_to = "Coverage_Level",
               values_to = "Coverage") %>%
  mutate(SD = case_when(
    Coverage_Level == "mean_coverage_50" ~ sd_coverage_50,
    Coverage_Level == "mean_coverage_80" ~ sd_coverage_80,
    Coverage_Level == "mean_coverage_95" ~ sd_coverage_95
  )) %>%
  mutate(
    lower = Coverage - 3 * SD,
    upper = Coverage + 3 * SD,
    Coverage_Level = recode(Coverage_Level, 
                            "mean_coverage_50" = "50%", 
                            "mean_coverage_80" = "80%", 
                            "mean_coverage_95" = "95%")
  )

ggplot(inter_coverage_plot_data, aes(x = Coverage_Level, y = Coverage, fill = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3, shape = 21, color = "black") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(width = 0.5)) +
  geom_hline(aes(yintercept = as.numeric(sub("%", "", Coverage_Level)) / 100), linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("mGP" = "skyblue", "tIWP2" = "salmon")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0.5, 0.8, 0.95)) +
  labs(title = "Interval Plot of Coverage Rates (Inter Scenario)",
       x = "Coverage Level",
       y = "Coverage Rate",
       fill = "Model") +
  theme_minimal()


# Prepare the data for the Pred scenario
pred_coverage_plot_data <- pred_summary %>%
  pivot_longer(cols = starts_with("mean_coverage"),
               names_to = "Coverage_Level",
               values_to = "Coverage") %>%
  mutate(SD = case_when(
    Coverage_Level == "mean_coverage_50" ~ sd_coverage_50,
    Coverage_Level == "mean_coverage_80" ~ sd_coverage_80,
    Coverage_Level == "mean_coverage_95" ~ sd_coverage_95
  )) %>%
  mutate(
    lower = Coverage - 3 * SD,
    upper = Coverage + 3 * SD,
    Coverage_Level = recode(Coverage_Level, 
                            "mean_coverage_50" = "50%", 
                            "mean_coverage_80" = "80%", 
                            "mean_coverage_95" = "95%")
  )

ggplot(pred_coverage_plot_data, aes(x = Coverage_Level, y = Coverage, fill = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3, shape = 21, color = "black") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(width = 0.5)) +
  geom_hline(aes(yintercept = as.numeric(sub("%", "", Coverage_Level)) / 100), linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("mGP" = "skyblue", "tIWP2" = "salmon")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0.5, 0.8, 0.95)) +
  labs(title = "Interval Plot of Coverage Rates (Pred Scenario)",
       x = "Coverage Level",
       y = "Coverage Rate",
       fill = "Model") +
  theme_minimal()
```







